---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# bccamtrap

<!-- badges: start -->
[![R-CMD-check](https://github.com/ateucher/bccamtrap/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/ateucher/bccamtrap/actions/workflows/R-CMD-check.yaml)
[![img](https://img.shields.io/badge/Lifecycle-Experimental-339999)](https://github.com/bcgov/repomountie/blob/master/doc/lifecycle-badges.md)
[![Codecov test coverage](https://codecov.io/gh/bcgov/bccamtrap/branch/main/graph/badge.svg)](https://app.codecov.io/gh/bcgov/bccamtrap?branch=main)
<!-- badges: end -->

Tools for QA and validation of Camera Trap data

## Installation

You can install the development version of bccamtrap from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("ateucher/bccamtrap")
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(bccamtrap)
## basic example code
```

### Project Status

### Getting Help or Reporting an Issue

To report bugs/issues/feature requests, please file an [issue](https://github.com/bcgov/bccamtrap/issues/).

### How to Contribute

If you would like to contribute, please see our [CONTRIBUTING](CONTRIBUTING.md) guidelines.

Please note that this project is released with a [Contributor Code of Conduct](CODE_OF_CONDUCT.md). By participating in this project you agree to abide by its terms.

### License

```
Copyright 2024 Province of British Columbia

Licensed under the Apache License, Version 2.0 (the &quot;License&quot;);
you may not use this file except in compliance with the License.
You may obtain a copy of the License at

http://www.apache.org/licenses/LICENSE-2.0

Unless required by applicable law or agreed to in writing, software distributed under the License is distributed on an &quot;AS IS&quot; BASIS,
WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or implied.
See the License for the specific language governing permissions and limitations under the License.
```
